NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2021-01-13 manuscriptPGPE.tex
Update figure 2
Updates from Overleaf

Here, we use natural abundance $^{13}$C NMR experiments and MD simulations from the NMRlipids open collaboration
to resolve the differences in conformational ensembles between PC, PE, PG and PS lipid headgroups.
Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
PE is also the second most abundant glycerophospholipid in eukaryotic cells
and has been related to various diseases \cite{vance15,calzada16,patel17}.
PS and PG are the most common negatively charged lipids in eukaryotes and bacteria, respectively,
and affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
All the studied lipids specifically bind to various proteins \cite{yeagle14}.
Furthermore, we use our results to elucidate lipid-protein interactions and the effect of charges on lipid conformations.

@@

Molecular dynamics simulation data were collected using
the NMRlipids Open Collaboration \cite{botan15}, which is running at
\url{nmrlipids.blogspot.fi} and \url{github.com/NMRlipids/NMRlipidsIVotherHGs}.
The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables \ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Tables~\ref{systemsMIX} and~\ref{systemsMIX2}.
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

\subsection{Analysis of molecular dynamics simulation data}
The large set of MD simulation data was analysed using the development version of NMRlipids databank.
Unique naming convention for lipid atoms in each force field was defined using the mapping files
and analysis for all simulations indexed in NMRlipids databank manner were performed
using python codes.

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python programs that use the MDAnalysis library \cite{agrawal11,gowers16}
used for all atom simulations is available in Ref. \citenum{MATCHgit}
({\tt scripts/calcOrderParameters.py}). For united atom simulations, the trajectories
with hydrogens having ideal geometry were constructed first using either buildH program~\cite{buildH}
or ({\tt scratch/opAAUA\_prod.py}) in  Ref. \citenum{MATCHgit}, and the order parameters were
then calculated from these trajectories. This approach has been tested against trajectories
with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
%\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

%\clearpage
\begin{figure*}[]
  \centering
%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  %  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
   \includegraphics[width=18.0cm]{./Figs/figure1.eps}
   \caption{\label{HGorderParameters}
     Chemical structure, refocused-INEPT spectrum, 2D R-PDLF spectra, and S-DROSS data (from top to bottom) of {\bf A)} POPE  and {\bf B)} POPG.
      Full NMR spectra are shown Figs. \ref{POPEspectra} and \ref{POPGspectra}.
     {\bf C)} Chemical structure of POPC and POPS.
    {\bf D)} Headgroup and glycerol backbone order parameters from different experiments in lamellar liquid disordered phase.
    The values and signs for POPE (310~K) and POPG (298~K)
    measured in this work, and for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}
    previously using $^{13}$C NMR. The literature values for
    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coliPE and E.coliPG (310~K) \cite{gally81}
    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
   }
   \todo{This is a sketch, Tiago Ferreira will make a new figure.} \\
%  \todo{D) could be clarified as Fig. 2 in the NMRlipids IVps paper.}
\end{figure*}

\subsection{Analysis of lipid conformations bound to proteins}
Lipid structures from PDB (\url{http://www.rcsb.org/})
were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api}).
First, all PDB entries containing PC, PE, PG, or PS lipid headgroups were searched.
The ligand names to identify the lipids were:
PLC, PX4, 6PL, LIO, HGX, PC7, PC8, P1O, 6O8, XP5, EGY, PLD, SBM, HXG, and PCW for PC;
8PE, PTY, 3PE, PEH, PEF, 6OE, 6O9, 9PE, PEV, 46E, SBJ, L9Q, PEK, EPH, ZPE, 9TL, 9Y0, 6OU, LOP, and PEE for PE;
PGT, PGK, LHG, 44G, PGV, OZ2, D3D, PGW, DR9, P6L, PG8, H3T, and GOT for PG; and
PSF, PS6, Q3G, P5S, D39, PS2, 17F, and 8SP for PS.
All PDB structures containing these ligands were downloaded and the dihedral angles of the first lipid in the latest version of the structures within each PDB code were calculated using the MDTraj python library \cite{mcgibbon15}.
The used Jupyter notebook is available from the project's GitHub repository ({\it scripts/pdbSEARCH.ipynb}).

@@

\begin{figure*}[bt]
  \centering
   \includegraphics[width=18.0cm]{./Figs/figure2.eps}
   \caption{\label{structures}
     Results from CHARMM36 simulations demonstrating the differences in conformational ensembles betwee different lipids. 
     {\bf A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
     {\bf B)} Headgroup and glycerol backbone region order parameters of different different lipids.
     {\bf C)} Distributions of heavy atom dihedral angles of different lipids from CHARMM36 simulations.
     {\bf D)} Distributions of P-N vector angle with respect to membrane normal.
  }
  \todo{This is a draft and requires quite a bit of polishing.
    More detailed discussion of this figure is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9} \\
\end{figure*}

@@

To experimentally characterize the headgroup conformational ensembles of lipids
that are not bound to proteins in electrostatically neutral cell membrane,
we measured the C-H bond order parameters and their signs from of POPG and POPE
in liquid lamellar phase, as we did previously for POPC and POPS \cite{ferreira13,ferreira16,antila19}.
Determination of headgroup and glycerol backbone order parameters and their signs
was straightforward from the data in Figs. \ref{HGorderParameters}, \ref{POPEspectra} and \ref{POPGspectra}
for all the C-H bonds, except the $\beta$ and g$_2$ carbons in POPG.
These carbons have overlapping peaks in the INEPT spectra due to their similar chemical environments,
and only the magnitude of the larger order parameter could be determined from the R-PDLF spectra (Fig. \ref{HGorderParameters} B).
Nevertheless, based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85},
we assigned the larger order parameter to the g$_2$ carbon
and used the literature value for the $\beta$-carbon in SIMPSON simulations to determine the signs.
The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive (Fig. \ref{HGorderParameters} (B)).
This interpretation is confirmed by SIMPSON calculations in Fig \ref{POPGsimpson}.

Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and headgroups from this and previous studies are collected in Fig. \ref{HGorderParameters} D), where signs determined from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature. Overall agreement of order parameters determined by different authors and different techniques for same lipid headgroup is very good in here and previous studies \cite{botan15,ollila16,antila19},
suggesting that the observed differences between lipid types arise from differences in headgroup chemistry rather than inaccuracies in experiments, or differences in acyl or experimental conditions. The most distinct order parameters are observed for PS headgroups, for which the $\alpha$-carbon order parameter exhibits significant forking and the $\beta$-carbon has more negative value than in other lipids. On the other hand, the $\beta$-carbon order parameter of PG headgroup has positive sign in contrast to all the other lipids. Notably, this has not been observed in traditional $^2$H NMR experiments, where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}. The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG. Essential differences between PC and PE headgroups are not observed.

@@

\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{./Figs/HGorderparametersPCvsSURFchangeDIHEDRALS.eps}
  \caption{\label{changesWITHsurf}
    {\bf A)} Modulation of PC headgroup order parameters and P-N vector angle upon addition of cationic surfactant
    from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
    {\bf B)} Changes in PC headgroup conformational ensembles upon increasing the amount of positive charge in bilayer,
    characterized by the heavy atom dihedral distributions, from CHARMM36 simulations.
  }
\end{figure*}

To resolve the conformational ensembles of lipid headgroups, we compared the headgroup and glycerol backbone C-H bond order parameters from different MD simulation force fields to the experimental data. None of the force fields reproduced the order parameters within experimental accuracy for any of the lipids analysed in this work (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}) or in our previous studies \cite{botan15,antila19}. Nevertheless, the CHARMM36 force field performs best for all the headgroups and reproduces the experimentally observed distinct order parameters of PS and PG lipids (Fig. \ref{structures} B). To understand the structural origin of these distinct order parameters, we calculated the distributions of heavy atom dihedral angles from CHARMM36 simulations in figure \ref{structures} C). Major differences between headgroups are observed only for the last two dihedrals in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$, which prefer {\it gauche$^-$} conformations for PG and {\it gauche$^+$} for PS. Rest of the dihedrals are similar between different lipids, with the exception of small differences for PS lipids. Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameter occur in the choline part, while also changes in phosphate region may contribute in PS lipids, which could also explain the more rigid headgroup structures~\cite{browning80,buldt81}.
Furthermore, the angle between headgroup dipole and membrane normal decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig. \ref{structures} B)). However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P-N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative in the CHARMM36 force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.

@@

Also binding of ions may affect the lipid headgroup conformational ensembles in physiological conditions.
In line with the results from cationic surfactants, 
the decreased probability for the trans state of g$_3$-O$_{g3}$-P-O$_\alpha$ dihedral is the main structural consequence of
bound ion to PC headgroup (Fig. \ref{DIHSwithCAlipid17eccPOPC} and \ref{DIHSwithCAcharmm36POPC}) in
Lipid17ecc (having the most realistic response of PC lipid headgroup order parameters to the calcium binding in Fig. \ref{changesWITHCaClPG})
and CHARMM36 simulations.
%simulations with the ECC included, where
%Similar but smaller effect is seen in CHARMM36 simulations with less bound calcium (Fig. \ref{DIHSwithCAcharmm36POPC}).
%Also the headgroup order parameters of
Response of charged lipids to the ion binding is less well understood and more difficult to capture in MD simulations \cite{antila19,melcr20}.
The most realistic simulations (Lipid17 and Slipids in figure \ref{changesWITHCaClPG}) suggest
%istic response of PG $\beta$-carbon to CaCl$_2$  even thought the calcium binding affinity was overestimated in these models.
%These models suggest
that the upward tilting of the headgroup dipole upon addition of CaCl$_2$ is weaker in PG than in PC headgroup,
%(\ref{changesWITHCaClPG}),
but dihedral distributions are more sensitive to the bound ions in PG (Figs \ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}).
However, the changes in dihedrals upon addition of calcium differ between the best models (Figs \ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}),
none of the simulations captures the calcium binding affinity and conformational ensemble of PG lipids simultaneously, and experimental data to evaluate the response of $\alpha$-carbon order parameters to the added calcium in PG is not available.
Therefore, more accurate force fields are required for the detailed analysis of the interactions between ions and charged lipids, as concluded previously also for PS lipids \cite{antila19,melcr20}.

@@

To analyze lipid conformations in protein bound states,
we calculated the heavy atom dihedral distributions from 
all lipid conformations available within protein structures
deposited in the PDB \cite{berman00} in figure \ref{dihedralsFROMpdb}.
We found 176 PC, 198 PE, 70 PG, and 41 PS conformations, which
present lipids that are tightly bound to proteins in fixed conformations
that can be determined as a part of protein structure using crystallography, NMR, or cryo-EM.

@@

Similarly to the liquid lamellar state, lipid dihedrals have very wide range of angles
when bound to proteins. Only %restrictions seem to be in 
P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P dihedrals
seem to avoid cis conformations as observed also in liquid lamellar state.
The %range of available angles for
O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and
g$_1$-g$_2$-g$_3$-O$_{g_3}$ dihedrals are even less restricted in protein bound state
than in liquid lamellar phase where cis and anti ecplipsed state were not present (Fig. \ref{structures}). 
Major differences between different lipid headgroups are not observed when bound to proteins.
Slight preference for the trans state in O$_\alpha$-C$_\alpha$-C$_\beta$-/C$_\gamma$ dihedral of PC,
and for positive angles in P-O$_\alpha$-C$_\alpha$-C$_\gamma$ dihedral of PS lipids with respect
to other lipids could be present in the data, but the statistics is not sufficient for conclusions.
Therefore, the differences in conformational ensembles between different
lipids in liquid lamellar state in Fig. \ref{structures} are not seen in the protein bound states.

This suggests that lipids can bind to proteins in wide range of conformations independently
of headgroup chemistry. Thereby, lipids compromize their preferred conformations when binding
to proteins and the binding of lipids to specific locations would be driven by 
the intermolecular interactions between proteins and lipids.
However, it is important to note that lipids are often not the main target in
structures deposited in PDB and therefore their conformations may be less reliable than proteins.
Stereochemical violations and structures deviating from lipid crystals have been
previously proposed to indicate inaccuracies in lipid structures in PDB \cite{marsh13b,pezeshkian18}. 
However, we see large deviations from lipid crystals structures also in conformational ensembles
that reproduce the NMR data in liquid lamellar phase, thereby proposing that such deviations
are realistic also in protein bound states.

@@

Conformational ensembles resolved using solid state NMR experiments and MD simulations
from NMRlipids open collaboration revealed that headgroups of the most abundant biological
phopholipids, PC, PE, PG and PS, sample wide range of different conformations in lamellar liquid state.
Differences in NMR order parameters between different headgroups can be explained
by the changes in dihedral angle distributions, suggesting that
similar conformations are accessed by all headgroups, but with different probabilities.
Also the changes in order parameters upon addition of charged molecules, such as cationic lipids, surfactants, or drugs,
in membranes can be explained by the changes in dihedral angle probabilities, particularly close to the phosphate region.

Wide range of conformations are observed also in lipids that are tightly bound to proteins in PDB,
suggesting that the specific binding of lipids to proteins is dominated by the intermolecular lipid-protein
interactions, while the differences in conformational ensembles between different lipid types
play a minor role. Therefore, lipids can conform themself to multiple binding positions locating
in various proteins.

Our results pave the way to understand how lipids regulate membrane protein function.
Lipid conformational ensembles in liquid lamellar phase are necessary
for the detailed analysis of lipid-protein interaction energetics, particularly for the entropic part. 
Furthermore, our results demonstrate how the NMRlipids databank, containing MD simulations and NMR data,
can be used to resolve conformational ensembles of disordered biomolecules in membrane environment,
and how this information can be coupled with structural data in the PDB.
We believe that this combination brings us closer to the comprehensive understanding of
complex biomolecular assemblies containg both disordered and structural regions, such as membrane
protein complexes.

>>>>

Here, we use natural abundance $^{13}$C NMR experiments and MD simulations from the NMRlipids open collaboration
to resolve the differences in conformational ensembles between PC, PE, PG and PS lipid headgroups.
Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
PE is also the second most abundant glycerophospholipid in eukaryotic cells
and has been related to various diseases \cite{vance15,calzada16,patel17}.
PS and PG are the most common negatively charged lipids in eukaryotes and bacteria, respectively,
and affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
All the studied lipids specifically bind to various proteins \cite{yeagle14}.
Furthermore, we use our results to elucidate lipid--protein interactions and the effect of charges on lipid conformations.

@@

Molecular dynamics simulation data were collected using
the NMRlipids Open Collaboration \cite{botan15}, which is running at
\url{nmrlipids.blogspot.fi} and \url{github.com/NMRlipids/NMRlipidsIVotherHGs}.
The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables~\ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Tables~\ref{systemsMIX} and \ref{systemsMIX2}.
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

\subsection{Analysis of molecular dynamics simulation data}
The large set of MD simulation data was analysed using the development version of NMRlipids databank.
Unique naming convention for lipid atoms in each force field was defined using the mapping files
and analysis for all simulations indexed in NMRlipids databank manner were performed
using Python codes.

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python programs that use the MDAnalysis library \cite{agrawal11,gowers16}
used for all atom simulations are available in Ref.~\citenum{MATCHgit}
(\texttt{scripts/calcOrderParameters.py}). For united atom simulations, the trajectories
with hydrogens having ideal geometry were constructed first using either \texttt{buildH} program~\cite{buildH}
or (\texttt{scratch/opAAUA\_prod.py}) in  Ref.~\citenum{MATCHgit}, and the order parameters were
then calculated from these trajectories. This approach has been tested against trajectories
with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
%\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\
The number density profiles of ions were calculated using the \texttt{gmx density} tool
of the GROMACS sofware package \cite{gromacsMANUAL}.

%\clearpage
\begin{figure*}[]
  \centering
%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  %  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
   \includegraphics[width=18.0cm]{./Figs/figure1.eps}
   \caption{\label{HGorderParameters}
     Chemical structure, refocused-INEPT spectrum, 2D R-PDLF spectra, and S-DROSS data (from top to bottom) of \textbf{A)} POPE  and \textbf{B)} POPG.
      Full NMR spectra are shown Figs. \ref{POPEspectra} and \ref{POPGspectra}.
     \textbf{C)} Chemical structure of POPC and POPS.
    \textbf{D)} Headgroup and glycerol backbone order parameters from different experiments in lamellar liquid disordered phase.
    The values and signs for POPE (310~K) and POPG (298~K)
    measured in this work, and for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}
    previously using $^{13}$C NMR. The literature values for
    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coliPE and E.coliPG (310~K) \cite{gally81}
    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
   }
   \todo{This is a sketch, Tiago Ferreira will make a new figure.} \\
%  \todo{D) could be clarified as Fig. 2 in the NMRlipids IVps paper.}
\end{figure*}

\subsection{Analysis of conformations of protein-bound lipids}
Lipid structures from Protein Data Bank (PDB, \url{http://www.rcsb.org/})
were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api}).
First, all PDB entries containing PC, PE, PG, or PS lipid headgroups were collected.
The ligand names to identify the lipids were:
PLC, PX4, 6PL, LIO, HGX, PC7, PC8, P1O, 6O8, XP5, EGY, PLD, SBM, HXG, and PCW for PC;
8PE, PTY, 3PE, PEH, PEF, 6OE, 6O9, 9PE, PEV, 46E, SBJ, L9Q, PEK, EPH, ZPE, 9TL, 9Y0, 6OU, LOP, and PEE for PE;
PGT, PGK, LHG, 44G, PGV, OZ2, D3D, PGW, DR9, P6L, PG8, H3T, and GOT for PG; and
PSF, PS6, Q3G, P5S, D39, PS2, 17F, and 8SP for PS.
Secondly, all PDB structures containing these ligands were downloaded and the dihedral angles of the first lipid in the latest version of the structures within each PDB code were calculated using the MDTraj python library \cite{mcgibbon15}.
The used Jupyter notebook is available from the project's GitHub repository (\textit{ scripts/pdbSEARCH.ipynb}).

@@

\begin{figure*}[bt]
  \centering
   \includegraphics[width=18.0cm]{./Figs/figure2.eps}
   \caption{\label{structures}
     Results from CHARMM36 simulations demonstrating the differences in conformational ensembles between different lipids. 
     \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
     \textbf{B)} Headgroup and glycerol backbone region order parameters of different different lipids.
     \textbf{C)} Distributions of heavy atom dihedral angles of different lipids from CHARMM36 simulations.
     \textbf{D)} Distributions of P-N vector angle with respect to membrane normal.
  }
  \todo{This is a draft and requires quite a bit of polishing.
    More detailed discussion of this figure is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9} \\
\end{figure*}

@@

To experimentally characterize the headgroup conformational ensembles of lipids
that are not bound to proteins in an electrostatically neutral cell membrane,
we measured the C--H bond order parameters and their signs from of POPG and POPE
in the liquid lamellar phase, as we did previously for POPC and POPS \cite{ferreira13,ferreira16,antila19}.
Determination of headgroup and glycerol backbone order parameters and their signs
was straightforward from the data in Figs.~\ref{HGorderParameters}, \ref{POPEspectra} and \ref{POPGspectra}
for all the C--H bonds, except for the $\beta$ and g$_2$ carbons in POPG.
These carbons have overlapping peaks in the INEPT spectra due to their similar chemical environments,
and only the magnitude of the larger order parameter could be determined from the R-PDLF spectra (Fig.~\ref{HGorderParameters}B)).
Nevertheless, based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85},
we assigned the larger order parameter to the g$_2$ carbon
and used the literature value for the $\beta$-carbon in SIMPSON simulations to determine the signs.
The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive (Fig.~\ref{HGorderParameters}B)).
This interpretation is confirmed by SIMPSON calculations in Fig.~\ref{POPGsimpson}.

Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and headgroups from this and previous studies are collected in Fig.~\ref{HGorderParameters}D), where signs determined from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature. The overall agreement of order parameters determined by different research teams and different techniques for the same lipid headgroup is very good here and in previous studies \cite{botan15,ollila16,antila19}. This suggests that the observed differences between lipid types arise from differences in headgroup chemistry rather than inaccuracies in experiments, or differences in the acyl chains or in experimental conditions. The most distinct order parameters are observed for PS headgroups, for which the $\alpha$-carbon order parameter exhibits significant forking and the $\beta$-carbon has more negative value than other studied lipid types. On the other hand, the $\beta$-carbon order parameter of PG headgroup has a positive sign, in contrast to all the other lipid types. Notably, this has not been observed in traditional $^2$H NMR experiments, where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}. The glycerol backbone order parameters are similar for all the lipid types, although they move slightly toward positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG. Essential differences between PC and PE headgroups are not observed.

@@

\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{./Figs/HGorderparametersPCvsSURFchangeDIHEDRALS.eps}
  \caption{\label{changesWITHsurf}
    \textbf{A)} Modulation of PC headgroup order parameters and P--N vector angle upon addition of cationic surfactant
    from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
    \textbf{B)} Changes in PC headgroup conformational ensembles upon increasing the amount of positive charge in bilayer,
    characterized by the heavy atom dihedral distributions, from CHARMM36 simulations.
  }
\end{figure*}

To resolve the conformational ensembles of lipid headgroups, we compared the headgroup and glycerol backbone C--H bond order parameters from different MD force fields to the experimental data. None of the force fields reproduced the order parameters within experimental accuracy for any of the lipids analysed in this work (Figs.~\ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}) or in our previous studies \cite{botan15,antila19}. Nevertheless, the CHARMM36 force field performs best for all the headgroups and reproduces the experimentally observed distinct order parameters of PS and PG lipids (Fig.~\ref{structures}B). To understand the structural origin of these distinct order parameters, we calculated the distributions of heavy atom dihedral angles from CHARMM36 simulations in Fig.~\ref{structures}C). Major differences between headgroups are observed only for the last two dihedrals in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$, which prefer \textit{gauche$^-$} conformations for PG and \textit{gauche$^+$} for PS. Rest of the dihedrals are similar between different lipids, with the exception of small differences for PS lipids. Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameter occur in the choline part, while also changes in phosphate region may contribute in PS lipids, which could also explain the more rigid headgroup structures~\cite{browning80,buldt81}.
Furthermore, the angle between headgroup dipole and membrane normal decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig.~\ref{structures}B)). However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative in the CHARMM36 force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.

@@

Also binding of ions may affect the lipid headgroup conformational ensembles in physiological conditions.
In line with the results from cationic surfactants, 
the decreased probability for the trans state of g$_3$-O$_{g3}$-P-O$_\alpha$ dihedral is the main structural consequence of
bound ion to PC headgroup (Fig.~\ref{DIHSwithCAlipid17eccPOPC} and \ref{DIHSwithCAcharmm36POPC}) in
Lipid17ecc (having the most realistic response of PC lipid headgroup order parameters to the calcium binding in Fig.~\ref{changesWITHCaClPG})
and CHARMM36 simulations.
%simulations with the ECC included, where
%Similar but smaller effect is seen in CHARMM36 simulations with less bound calcium (Fig. \ref{DIHSwithCAcharmm36POPC}).
%Also the headgroup order parameters of
Response of charged lipids to the ion binding is less well understood and more difficult to capture in MD simulations \cite{antila19,melcr20}.
The most realistic simulations (Lipid17 and Slipids in Fig.~\ref{changesWITHCaClPG}) suggest
%istic response of PG $\beta$-carbon to CaCl$_2$  even thought the calcium binding affinity was overestimated in these models.
%These models suggest
that the upward tilting of the headgroup dipole upon addition of CaCl$_2$ is weaker in PG than in PC headgroup,
%(\ref{changesWITHCaClPG}),
but dihedral distributions are more sensitive to the bound ions in PG (Figs.~\ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}).
However, the changes in dihedrals upon addition of calcium differ between the best models (Figs.~\ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}),
none of the simulations captures the calcium binding affinity and conformational ensemble of PG lipids simultaneously, and experimental data to evaluate the response of $\alpha$-carbon order parameters to the added calcium in PG is not available.
Therefore, more accurate force fields are required for the detailed analysis of the interactions between ions and charged lipids, as concluded previously also for PS lipids \cite{antila19,melcr20}.

@@

To analyze lipid conformations in protein bound states,
we calculated the heavy atom dihedral distributions from 
all lipid conformations available within protein structures
deposited in the PDB \cite{berman00} in Fig.~\ref{dihedralsFROMpdb}.
We found 176 PC, 198 PE, 70 PG, and 41 PS conformations, which
present lipids that are tightly bound to proteins in fixed conformations
that can be determined as a part of protein structure using crystallography, NMR, or cryo-EM.

@@

Similarly to the liquid lamellar state, lipid dihedrals have very wide range of angles
when bound to proteins. Only %restrictions seem to be in 
P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P dihedrals
seem to avoid $cis$ conformations as observed also in liquid lamellar state.
The %range of available angles for
O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and
g$_1$-g$_2$-g$_3$-O$_{g_3}$ dihedrals are even less restricted in protein bound state
than in liquid lamellar phase where $cis$ and $anti$ eclipsed state were not present (Fig.~\ref{structures}). 
Major differences between different lipid headgroups are not observed when bound to proteins.
Slight preference for the trans state in O$_\alpha$-C$_\alpha$-C$_\beta$-/C$_\gamma$ dihedral of PC,
and for positive angles in P-O$_\alpha$-C$_\alpha$-C$_\gamma$ dihedral of PS lipids with respect
to other lipids could be present in the data, but the statistics is not sufficient for conclusions.
Therefore, the differences in conformational ensembles between different
lipids in liquid lamellar state in Fig.~\ref{structures} are not seen in the protein bound states.

This suggests that lipids can bind to proteins in wide range of conformations independently
of headgroup chemistry. Thereby, lipids compromise their preferred conformations when binding
to proteins and the binding of lipids to specific locations would be driven by 
the intermolecular interactions between proteins and lipids.
However, it is important to note that lipids are often not the main target in
structures deposited in PDB and therefore their conformations may be less reliable than those of proteins.
Stereochemical violations and structures deviating from lipid crystals have been
previously proposed to indicate inaccuracies in lipid structures in PDB \cite{marsh13b,pezeshkian18}. 
However, we see large deviations from lipid crystals structures also in conformational ensembles
that reproduce the NMR data in liquid lamellar phase, thereby proposing that such deviations
are realistic also in protein-bound states.

@@

Conformational ensembles resolved using solid state NMR experiments and MD simulations
from NMRlipids open collaboration revealed that headgroups of the most abundant biological
phopholipids, PC, PE, PG and PS, sample a wide range of different conformations in the lamellar liquid state.
Differences in NMR order parameters between different headgroups can be explained
by the changes in dihedral angle distributions, suggesting that
similar conformations are accessed by all headgroups, but with different probabilities.
Also the changes in order parameters upon the addition of charged molecules---such as cationic lipids, surfactants, or drugs---in membranes can be explained by the changes in dihedral angle probabilities, particularly close to the phosphate region.

Wide range of conformations are observed also in lipids that are tightly bound to proteins in PDB,
suggesting that the specific binding of lipids to proteins is dominated by the intermolecular lipid--protein
interactions, while the differences in conformational ensembles between different lipid types
play a minor role. Therefore, lipids can conform themselves to multiple binding positions located
in various proteins.

Our results pave the way to understanding how lipids regulate membrane protein function.
Lipid conformational ensembles in liquid lamellar phase are necessary
for the detailed analysis of lipid--protein interaction energetics, particularly for its entropic component. 
Furthermore, our results demonstrate how the NMRlipids databank, containing MD simulations and NMR data,
can be used to resolve conformational ensembles of disordered biomolecules in a membrane environment,
and how this information can be coupled with structural data in the PDB.
We believe that this combination brings us closer to the comprehensive understanding of
complex biomolecular assemblies containing both disordered and structural regions, such as membrane
protein complexes.
